---
title: "Elo Ratings"
output: html_document
editor_options: 
  chunk_output_type: console
---

Elo ratings: Why is it powerful? It automatically adjusts for strength of opponent: beating top players moves you up fast, beating weak players not much. Gives you a single number to rank players or teams, but it implicitly accounts for all games played. Great for dynamic tracking: you can plot Elo vs time to see if someone’s getting better or declining.

The **Elo rating system** is a way to estimate the skill levels of players in competitive games.

It was invented by **Arpad Elo**, a Hungarian-American physics professor and chess master.\
The system became widely used in chess, but is now used for many sports and games, including online gaming and even evaluating football teams.

The system works like this:

-   Each player starts with an initial rating (commonly 1500).
-   When two players (or teams) play a match:
    -   The expected outcome is calculated based on the difference in their ratings.
    -   If a player performs better than expected (wins against a stronger opponent), they gain more points.
    -   If a player performs worse than expected (loses to a weaker opponent), they lose more points.
-   The **K-factor** controls how much the ratings adjust after each game.

The expected score (probability of winning) is:

$$
E_A = \frac{1}{1 + 10^{(R_B - R_A) / 400}}
$$

where: - $R_A$ is Player A’s rating - $R_B$ is Player B’s rating

Then after the match:

$$
R'_A = R_A + K \times (S_A - E_A)
$$

where: $S_A = 1$ if A wins, $0.5$ if draw, $0$ if A loses and $K$ is the adjustment factor (commonly 32).

✅ It’s **self-correcting**:\
If a player wins consistently, their rating rises, so future wins give less of a boost (and losses hurt more).

✅ It’s **continuous**:\
It tracks performance game by game, without needing to wait for a whole season.

✅ It’s **predictive**:\
A higher rating difference means a higher expected probability of winning.

-   Players with the same rating are expected to win about equally often.
-   The ratings adjust automatically to how surprising the outcomes are.
-   It can handle long histories of matchups without starting over.

You’re using this system to track and compare players like **Jim Elsner**, **Mike Prost**, and others across multiple tournaments.

By plotting Elo over time, you can see:

-   who’s improving
-   who’s consistent
-   who might be having up or down seasons.

## References

-   Arpad Elo, *The Rating of Chessplayers, Past and Present*, 1978.
-   FIDE Handbook on Elo calculation: <https://handbook.fide.com/chapter/B022017>
-   Wikipedia: <https://en.wikipedia.org/wiki/Elo_rating_system>

```{r}
library(dplyr)
library(tidyr)

# Load data
df <- read.csv("data/LeagueGames.csv", stringsAsFactors = FALSE) 

# Create a data frame to store win/loss for each player
player_records <- data.frame(Player = character(), Wins = integer(), Losses = integer(), stringsAsFactors = FALSE)

record_win <- function(player) {
  if (!(player %in% player_records$Player)) {
    player_records <<- rbind(player_records, data.frame(Player = player, Wins = 0, Losses = 0))
  }
  player_records[player_records$Player == player, "Wins"] <<- player_records[player_records$Player == player, "Wins"] + 1
}

record_loss <- function(player) {
  if (!(player %in% player_records$Player)) {
    player_records <<- rbind(player_records, data.frame(Player = player, Wins = 0, Losses = 0))
  }
  player_records[player_records$Player == player, "Losses"] <<- player_records[player_records$Player == player, "Losses"] + 1
}

# Initialize Elo ratings (default to 1500)
player_elos <- new.env()
default_elo <- 1500
k_factor <- 32

get_elo <- function(player) {
  if (!exists(player, envir = player_elos)) {
    assign(player, default_elo, envir = player_elos)
  }
  get(player, envir = player_elos)
}

set_elo <- function(player, new_rating) {
  assign(player, new_rating, envir = player_elos)
}

update_elo <- function(player_a, player_b, outcome, k = k_factor) {
  rating_a <- get_elo(player_a)
  rating_b <- get_elo(player_b)
  expected_a <- 1 / (1 + 10^((rating_b - rating_a) / 400))
  rating_a_new <- rating_a + k * (outcome - expected_a)
  rating_b_new <- rating_b + k * ((1 - outcome) - (1 - expected_a))
  set_elo(player_a, rating_a_new)
  set_elo(player_b, rating_b_new)
}

# Process each game
for (i in 1:nrow(df)) {
  A1 <- df$`Team.A.Player.1`[i]
  A2 <- df$`Team.A.Player.2`[i]
  B1 <- df$`Team.B.Player.1`[i]
  B2 <- df$`Team.B.Player.2`[i]
  score_a <- df$Score.A[i]
  score_b <- df$Score.B[i]

  elo_A <- (get_elo(A1) + get_elo(A2)) / 2
  elo_B <- (get_elo(B1) + get_elo(B2)) / 2

  expected_A <- 1 / (1 + 10^((elo_B - elo_A) / 400))
  outcome_A <- ifelse(score_a > score_b, 1, 0)
  delta <- k_factor * (outcome_A - expected_A)

  for (player in c(A1, A2)) {
    set_elo(player, get_elo(player) + delta / 2)
    if (outcome_A == 1) {
      record_win(player)
    } else {
      record_loss(player)
    }
  }
  for (player in c(B1, B2)) {
    set_elo(player, get_elo(player) - delta / 2)
    if (outcome_A == 0) {
      record_win(player)
    } else {
      record_loss(player)
    }
  }
}

# Combine Elo and win-loss record
final_elos <- as.data.frame(as.list(player_elos)) %>%
  tidyr::pivot_longer(cols = everything(), names_to = "Player", values_to = "Elo") %>%
  left_join(player_records, by = "Player") %>%
  mutate(Wins = replace_na(Wins, 0),
         Losses = replace_na(Losses, 0),
         WinPct = round(Wins / (Wins + Losses), 3),
         Games = Wins + Losses) %>%
  filter(Games >= 25) %>%
  arrange(desc(Elo))

print(final_elos, n = 25)
```

Script with Elo history tracking & plotting by ordinal game count
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Load and preprocess
df <- read.csv("data/LeagueGames.csv", stringsAsFactors = FALSE)
df$Date <- as.Date(df$Date)
df <- df[order(df$Date), ]  # Ensure games are processed in order

# Initialize
player_records <- data.frame(Player = character(), Wins = integer(), Losses = integer(), stringsAsFactors = FALSE)
player_elos <- new.env()
elo_history <- data.frame(GameNumber = integer(), Player = character(), Elo = numeric(), stringsAsFactors = FALSE)
default_elo <- 1500
k_factor <- 32
game_counter <- 1

get_elo <- function(player) {
  if (!exists(player, envir = player_elos)) {
    assign(player, default_elo, envir = player_elos)
  }
  get(player, envir = player_elos)
}

set_elo <- function(player, new_rating) {
  assign(player, new_rating, envir = player_elos)
}

record_win <- function(player) {
  if (!(player %in% player_records$Player)) {
    player_records <<- rbind(player_records, data.frame(Player = player, Wins = 0, Losses = 0))
  }
  player_records[player_records$Player == player, "Wins"] <<- player_records[player_records$Player == player, "Wins"] + 1
}

record_loss <- function(player) {
  if (!(player %in% player_records$Player)) {
    player_records <<- rbind(player_records, data.frame(Player = player, Wins = 0, Losses = 0))
  }
  player_records[player_records$Player == player, "Losses"] <<- player_records[player_records$Player == player, "Losses"] + 1
}

# Process games
for (i in 1:nrow(df)) {
  A1 <- df$`Team.A.Player.1`[i]
  A2 <- df$`Team.A.Player.2`[i]
  B1 <- df$`Team.B.Player.1`[i]
  B2 <- df$`Team.B.Player.2`[i]
  score_a <- df$Score.A[i]
  score_b <- df$Score.B[i]

  elo_A <- (get_elo(A1) + get_elo(A2)) / 2
  elo_B <- (get_elo(B1) + get_elo(B2)) / 2

  expected_A <- 1 / (1 + 10^((elo_B - elo_A) / 400))
  outcome_A <- ifelse(score_a > score_b, 1, 0)
  delta <- k_factor * (outcome_A - expected_A)

  for (player in c(A1, A2)) {
    new_elo <- get_elo(player) + delta / 2
    set_elo(player, new_elo)
    if (outcome_A == 1) {
      record_win(player)
    } else {
      record_loss(player)
    }
    elo_history <- rbind(elo_history, data.frame(GameNumber = game_counter, Player = player, Elo = new_elo))
  }

  for (player in c(B1, B2)) {
    new_elo <- get_elo(player) - delta / 2
    set_elo(player, new_elo)
    if (outcome_A == 0) {
      record_win(player)
    } else {
      record_loss(player)
    }
    elo_history <- rbind(elo_history, data.frame(GameNumber = game_counter, Player = player, Elo = new_elo))
  }

  game_counter <- game_counter + 1
}

# Plot Elo trajectory using game number
plot_elo_ordinal <- function(player_name) {
  elo_history %>%
    filter(Player == player_name) %>%
    ggplot(aes(x = GameNumber, y = Elo)) +
    geom_line(color = "steelblue", linewidth = 1) +
    geom_point(size = 2) +
    labs(title = paste("Elo Rating Over Games for", player_name),
         x = "Game Number (Ordinal)",
         y = "Elo Rating") +
    theme_minimal()
}

# Example:
plot_elo_ordinal("Drew")

# Plot Elo over ordinal games for multiple players
plot_elo_multi <- function(player_names) {
  elo_history %>%
    filter(Player %in% player_names) %>%
    ggplot(aes(x = GameNumber, y = Elo, color = Player)) +
    geom_line(linewidth = 1.2) +
    labs(title = "Elo Ratings Over Games",
         x = "Game Number (Ordinal)",
         y = "Elo Rating",
         color = "Player") +
    theme_minimal() +
    theme(legend.position = "right")
}

# Example usage: compare 4 players
plot_elo_multi(c("Jim", "Jacob", "Craig", "Jordan", "Drew", "JaSun"))

```

Maya's cumultative point differential over time

```{r}
library(dplyr)
library(ggplot2)

# Step 1: Sort games by date
df$Date <- as.Date(df$Date)
df <- df[order(df$Date), ]
df$GameNumber <- 1:nrow(df)

# Step 2: Identify games Maya played in and calculate point differential
maya_diff <- df %>%
  filter(`Team.A.Player.1` == "Maya" | `Team.A.Player.2` == "Maya" |
         `Team.B.Player.1` == "Maya" | `Team.B.Player.2` == "Maya") %>%
  mutate(
    Team = ifelse(`Team.A.Player.1` == "Maya" | `Team.A.Player.2` == "Maya", "A", "B"),
    Diff = ifelse(Team == "A", Score.A - Score.B, Score.B - Score.A)
  ) %>%
  select(GameNumber, Date, Diff) %>%
  arrange(GameNumber) %>%
  mutate(CumDiff = cumsum(Diff))

# Step 3: Plot
ggplot(maya_diff, aes(x = GameNumber, y = CumDiff)) +
  geom_line(color = "firebrick", size = 1.2) +
  labs(title = "Maya's Cumulative Point Differential Over Time",
       x = "Game Number (Ordinal)",
       y = "Cumulative Point Differential") +
  theme_minimal()

```

Script with Elo history tracking & plotting by date
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Load data
df <- read.csv("data/LeagueGames.csv", stringsAsFactors = FALSE)
df$Date <- as.Date(df$Date)

# Create win/loss record
player_records <- data.frame(Player = character(), Wins = integer(), Losses = integer(), stringsAsFactors = FALSE)

record_win <- function(player) {
  if (!(player %in% player_records$Player)) {
    player_records <<- rbind(player_records, data.frame(Player = player, Wins = 0, Losses = 0))
  }
  player_records[player_records$Player == player, "Wins"] <<- player_records[player_records$Player == player, "Wins"] + 1
}

record_loss <- function(player) {
  if (!(player %in% player_records$Player)) {
    player_records <<- rbind(player_records, data.frame(Player = player, Wins = 0, Losses = 0))
  }
  player_records[player_records$Player == player, "Losses"] <<- player_records[player_records$Player == player, "Losses"] + 1
}

# Initialize Elo environment
player_elos <- new.env()
default_elo <- 1500
k_factor <- 32

get_elo <- function(player) {
  if (!exists(player, envir = player_elos)) {
    assign(player, default_elo, envir = player_elos)
  }
  get(player, envir = player_elos)
}

set_elo <- function(player, new_rating) {
  assign(player, new_rating, envir = player_elos)
}

# Initialize Elo history tracking
elo_history <- data.frame(Date = as.Date(character()),
                          Player = character(),
                          Elo = numeric(),
                          stringsAsFactors = FALSE)

# Process each game and record Elo history
df <- df[order(df$Date), ]

for (i in 1:nrow(df)) {
  row <- df[i, ]
  A1 <- row$Team.A.Player.1
  A2 <- row$Team.A.Player.2
  B1 <- row$Team.B.Player.1
  B2 <- row$Team.B.Player.2
  score_a <- row$Score.A
  score_b <- row$Score.B
  game_date <- row$Date

  elo_A <- (get_elo(A1) + get_elo(A2)) / 2
  elo_B <- (get_elo(B1) + get_elo(B2)) / 2

  expected_A <- 1 / (1 + 10^((elo_B - elo_A) / 400))
  outcome_A <- ifelse(score_a > score_b, 1, 0)
  delta <- k_factor * (outcome_A - expected_A)

  for (player in c(A1, A2)) {
    new_elo <- get_elo(player) + delta / 2
    set_elo(player, new_elo)
    if (outcome_A == 1) {
      record_win(player)
    } else {
      record_loss(player)
    }
    elo_history <- rbind(elo_history, data.frame(Date = game_date, Player = player, Elo = new_elo))
  }

  for (player in c(B1, B2)) {
    new_elo <- get_elo(player) - delta / 2
    set_elo(player, new_elo)
    if (outcome_A == 0) {
      record_win(player)
    } else {
      record_loss(player)
    }
    elo_history <- rbind(elo_history, data.frame(Date = game_date, Player = player, Elo = new_elo))
  }
}

# Plot Elo rating for a specific player
plot_elo <- function(player_name) {
  p_data <- elo_history %>%
    filter(Player == player_name) %>%
    arrange(Date)

  ggplot(p_data, aes(x = Date, y = Elo)) +
    geom_line(color = "steelblue", linewidth = 1) +
    geom_point(size = 2) +
    labs(title = paste("Elo Rating Over Time for", player_name),
         x = "Date", y = "Elo Rating") +
    theme_minimal()
}

# Example usage:
plot_elo("Maya")

```

Network of successful partnerships
```{r}
library(dplyr)
library(tidyr)
library(igraph)
library(ggraph)

# Load and sort game data
df <- read.csv("data/LeagueGames.csv", stringsAsFactors = FALSE)
df$Date <- as.Date(df$Date)
df <- df[order(df$Date), ]

# Store win/loss records between teammates
teammate_records <- list()

`%||%` <- function(a, b) if (!is.null(a)) a else b  # safe default operator

for (i in 1:nrow(df)) {
  row <- df[i, ]
  team_a <- c(row$Team.A.Player.1, row$Team.A.Player.2)
  team_b <- c(row$Team.B.Player.1, row$Team.B.Player.2)
  winner <- ifelse(row$Score.A > row$Score.B, "A", "B")

  team_a_key <- paste(sort(team_a), collapse = "_")
  teammate_records[[team_a_key]] <- teammate_records[[team_a_key]] %||% list(W = 0, L = 0)
  if (winner == "A") {
    teammate_records[[team_a_key]]$W <- teammate_records[[team_a_key]]$W + 1
  } else {
    teammate_records[[team_a_key]]$L <- teammate_records[[team_a_key]]$L + 1
  }

  team_b_key <- paste(sort(team_b), collapse = "_")
  teammate_records[[team_b_key]] <- teammate_records[[team_b_key]] %||% list(W = 0, L = 0)
  if (winner == "B") {
    teammate_records[[team_b_key]]$W <- teammate_records[[team_b_key]]$W + 1
  } else {
    teammate_records[[team_b_key]]$L <- teammate_records[[team_b_key]]$L + 1
  }
}

# Convert teammate win-loss data to a data frame
teammate_df <- tibble::tibble()
for (pair in names(teammate_records)) {
  players <- unlist(strsplit(pair, "_"))
  W <- teammate_records[[pair]]$W
  L <- teammate_records[[pair]]$L
  G <- W + L
  WLpct <- ifelse(G > 0, W / G, NA)
  teammate_df <- bind_rows(teammate_df,
                           data.frame(Player1 = players[1], Player2 = players[2],
                                      Wins = W, Losses = L, Games = G, WLpct = WLpct))
}

# Only keep teammate pairs with >5 games together
teammate_df <- teammate_df %>%
  filter(Games >= 3)

# Find best teammate (highest W-L %) for each player
best_partners <- teammate_df %>%
  mutate(PlayerA = Player1, PlayerB = Player2) %>%
  bind_rows(
    select(., Player = PlayerA, Partner = PlayerB, Games, WLpct),
    select(., Player = PlayerB, Partner = PlayerA, Games, WLpct)
  ) %>%
  group_by(Player) %>%
  filter(Games > 5) %>%
  arrange(desc(WLpct)) %>%
  slice(1) %>%
  ungroup()

# Create directed edges
edges <- best_partners %>%
  select(from = Player, to = Partner, weight = WLpct) %>%
  mutate(pair_id = paste(pmin(from, to), pmax(from, to), sep = "_"))

# Identify bidirectional edges
bidirectional <- edges %>%
  group_by(pair_id) %>%
  filter(n() == 2) %>%
  ungroup()

edges <- edges %>%
  mutate(bidirectional = pair_id %in% bidirectional$pair_id)

# Create igraph object
g <- graph_from_data_frame(edges, directed = TRUE)

# Plot with ggraph
g <- delete_vertices(g, V(g)[name == "NA"])
ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = weight),
                 arrow = arrow(length = unit(4, "mm"), type = "closed"),
                 end_cap = circle(3, 'mm'),
                 start_cap = circle(3, 'mm'),
                 color = "gray80",
                 lineend = "round") +
  scale_edge_width(range = c(0.5, 2.5), guide = "none") +
  geom_node_point(size = 5, color = "steelblue") +
  geom_node_text(aes(label = name), repel = TRUE, size = 4) +
  theme_void() +
  labs(title = "Best Teammate Network")
```

Teammate W–L % ranked for each player
```{r}
library(dplyr)
library(tidyr)

# Load game data
df <- read.csv("data/LeagueGames.csv", stringsAsFactors = FALSE)

# Step 1: build team records by sorted pair of players
team_records <- list()

for (i in 1:nrow(df)) {
  row <- df[i, ]
  
  # Define teams
  team_a <- sort(c(row$Team.A.Player.1, row$Team.A.Player.2))
  team_b <- sort(c(row$Team.B.Player.1, row$Team.B.Player.2))
  winner <- ifelse(row$Score.A > row$Score.B, "A", "B")

  # A team record
  key_a <- paste(team_a, collapse = "_")
  if (!key_a %in% names(team_records)) {
    team_records[[key_a]] <- list(players = team_a, wins = 0, losses = 0)
  }
  if (winner == "A") {
    team_records[[key_a]]$wins <- team_records[[key_a]]$wins + 1
  } else {
    team_records[[key_a]]$losses <- team_records[[key_a]]$losses + 1
  }

  # B team record
  key_b <- paste(team_b, collapse = "_")
  if (!key_b %in% names(team_records)) {
    team_records[[key_b]] <- list(players = team_b, wins = 0, losses = 0)
  }
  if (winner == "B") {
    team_records[[key_b]]$wins <- team_records[[key_b]]$wins + 1
  } else {
    team_records[[key_b]]$losses <- team_records[[key_b]]$losses + 1
  }
}

# Step 2: expand to player-wise view
player_teammate_stats <- data.frame(Player = character(), 
                                    Teammate = character(), 
                                    Wins = integer(), 
                                    Losses = integer(),
                                    WLpct = numeric(),
                                    stringsAsFactors = FALSE)

for (record in team_records) {
  p1 <- record$players[1]
  p2 <- record$players[2]
  wins <- record$wins
  losses <- record$losses
  total <- wins + losses
  wlpct <- ifelse(total > 0, wins / total, NA)

  player_teammate_stats <- bind_rows(
    player_teammate_stats,
    data.frame(Player = p1, Teammate = p2, Wins = wins, Losses = losses, WLpct = round(wlpct, 3)),
    data.frame(Player = p2, Teammate = p1, Wins = wins, Losses = losses, WLpct = round(wlpct, 3))
  )
}

# Step 3: sort teammates for each player by WL %
ranked_teammates <- player_teammate_stats %>%
  filter(Wins + Losses > 5) %>%
  arrange(Player, WLpct)

# View result
print(ranked_teammates)

```




OLDER
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(PlayerRatings)

games_df <- read_csv(here::here("data", "LeagueGames.csv")) |>
  filter(Type == "Regular")

# Create all winner vs loser records
pairwise_results <- games_df %>%
  filter(!is.na(`Score A`) & !is.na(`Score B`)) %>%
  mutate(
    Winning_P1 = if_else(`Score A` > `Score B`, `Team A Player 1`, `Team B Player 1`),
    Winning_P2 = if_else(`Score A` > `Score B`, `Team A Player 2`, `Team B Player 2`),
    Losing_P1  = if_else(`Score A` < `Score B`, `Team A Player 1`, `Team B Player 1`),
    Losing_P2  = if_else(`Score A` < `Score B`, `Team A Player 2`, `Team B Player 2`)
  ) %>%
  select(Winning_P1, Winning_P2, Losing_P1, Losing_P2) %>%
  pivot_longer(cols = c(Winning_P1, Winning_P2), names_to = "win_role", values_to = "Winner") %>%
  pivot_longer(cols = c(Losing_P1, Losing_P2), names_to = "lose_role", values_to = "Loser") %>%
  transmute(Date = 1, Player1 = Winner, Player2 = Loser, Score = 1) %>%
  distinct()

# Now add inverse games (losses count as 0)
elo_input <- bind_rows(
  pairwise_results,
  pairwise_results %>%
    mutate(Player1 = Player2, Player2 = Player1, Score = 0)
)

# Run Elo model
elo_ratings <- elo(elo_input, init = 1500, k = 32)

# Show top 25 players
elo_ratings$ratings %>%
  arrange(desc(Rating)) %>%
  mutate(Rating = round(Rating, 0)) %>%
  slice_head(n = 25)
```

Elo Escape!! player ratings over time

```{r}
library(readr)
# Read your data
games_df <- read_csv(here("data", "TallyRallyGames.csv"))
```

# Create proper Date and numeric time index

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(PlayerRatings)
library(here)

games_df <- games_df %>%
  mutate(
    Month_Day_Clean = str_remove(`Month Day`, "(st|nd|rd|th)"),
    Date_str = paste(Month_Day_Clean, Year),
    Date = mdy(Date_str)
  )

# Create numeric index (e.g. days since first date)
games_df <- games_df %>%
  arrange(Date) %>%
  mutate(Date_num = as.integer(difftime(Date, min(Date, na.rm = TRUE), units = "days")))

# Build pairwise winner-loser data
pairwise_results <- games_df %>%
  filter(!is.na(`Score A`) & !is.na(`Score B`)) %>%
  mutate(
    Winning_P1 = if_else(`Score A` > `Score B`, `Team A Player 1`, `Team B Player 1`),
    Winning_P2 = if_else(`Score A` > `Score B`, `Team A Player 2`, `Team B Player 2`),
    Losing_P1  = if_else(`Score A` < `Score B`, `Team A Player 1`, `Team B Player 1`),
    Losing_P2  = if_else(`Score A` < `Score B`, `Team A Player 2`, `Team B Player 2`)
  ) %>%
  select(Date_num, Winning_P1, Winning_P2, Losing_P1, Losing_P2) %>%
  pivot_longer(c(Winning_P1, Winning_P2), names_to = "win_role", values_to = "Winner") %>%
  pivot_longer(c(Losing_P1, Losing_P2), names_to = "lose_role", values_to = "Loser") %>%
  transmute(Date = Date_num, Player1 = Winner, Player2 = Loser, Score = 1) %>%
  distinct()

# Build input for Elo (losses as 0)
elo_input <- bind_rows(
  pairwise_results,
  pairwise_results %>%
    mutate(Player1 = Player2, Player2 = Player1, Score = 0)
)

# Run Elo model
elo_run <- elo(elo_input, init = 1500, k = 32, history = TRUE)

#  Print final ratings
elo_ratings$ratings %>%
  arrange(desc(Rating)) %>%
  mutate(Rating = round(Rating, 0)) %>%
  slice_head(n = 25)
```

```{r}
library(purrr)

# Extract dim names
players <- dimnames(elo_run$history)[[1]]
periods <- dimnames(elo_run$history)[[2]]

# Build tidy data frame
elo_history <- map_dfr(seq_along(periods), function(i) {
  tibble(
    Period = as.integer(periods[i]),
    Player = players,
    Rating = elo_run$history[, i, "Rating"]
  )
})

# Filter to your players
players_to_plot <- c("Jim Elsner", "Diana Elsner", "JaSun Burdick", "Craig Anderson", "Hank Bass", "Drew Magee", "Jordan Huston", "Scott Schmucker", "Conrad Damon", "Harvey Brandt", "John Elsner", "Mike Prost")
elo_history_filtered <- elo_history %>%
  filter(Player %in% players_to_plot)

library(lubridate)

# Reconstruct dates from Date_num
period_dates <- elo_input %>%
  distinct(Date) %>%
  arrange(Date) %>%
  mutate(Period = row_number(),
         Actual_Date = min(games_df$Date, na.rm = TRUE) + days(Date)) %>%
  select(Period, Actual_Date)

elo_history_with_date <- elo_history %>%
  left_join(period_dates, by = "Period")

ggplot(elo_history_with_date %>% filter(Player %in% players_to_plot),
       aes(x = Actual_Date, y = Rating, color = Player)) +
  geom_line(alpha = 0.7) +  # main lines
  geom_smooth(se = FALSE, method = "loess", span = 0.75) +  # smoothed trend line
  facet_wrap(~ Player, scales = "fixed", ncol = 3) +  # small multiples
  labs(
    title = "Elo Ratings Over Time by Player",
    x = "Date",
    y = "Elo Rating"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Elo Escape!! team ratings over time

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(PlayerRatings)
library(lubridate)

games_df <- read_csv(here::here("data", "TallyRallyGames.csv"))

# 1. Prepare data
games_df <- games_df %>%
  mutate(
    # Clean and construct Date
    Month_Day_Clean = str_remove(`Month Day`, "(st|nd|rd|th)"),
    Date_str = paste(Month_Day_Clean, Year),
    Date = mdy(Date_str)
  ) %>%
  arrange(Date) %>%
  mutate(Date_num = as.integer(difftime(Date, min(Date, na.rm = TRUE), units = "days")))

# 2. Build team labels (sorted so "Jim Elsner & John Elsner" always appears that way)
games_df <- games_df %>%
  mutate(
    Team_A = map2_chr(`Team A Player 1`, `Team A Player 2`, ~ paste(sort(c(.x, .y)), collapse = " & ")),
    Team_B = map2_chr(`Team B Player 1`, `Team B Player 2`, ~ paste(sort(c(.x, .y)), collapse = " & "))
  )

# 3. Create pairwise results for Elo
team_results <- games_df %>%
  filter(!is.na(`Score A`) & !is.na(`Score B`)) %>%
  transmute(
    Date = Date_num,
    Team1 = if_else(`Score A` > `Score B`, Team_A, Team_B),
    Team2 = if_else(`Score A` > `Score B`, Team_B, Team_A),
    Score = 1
  )

# Add the inverse games (losses count as 0)
team_elo_input <- bind_rows(
  team_results,
  team_results %>%
    transmute(Date = Date, Team1 = Team2, Team2 = Team1, Score = 0)
)

# 4. Run Elo model
team_elo_run <- elo(team_elo_input, init = 1500, k = 32, history = TRUE)
```

Plot
```{r}
library(purrr)
library(ggplot2)

# Extract team names and periods
teams <- dimnames(team_elo_run$history)[[1]]
periods <- dimnames(team_elo_run$history)[[2]]

# Build tidy data
team_elo_history <- map_dfr(seq_along(periods), function(i) {
  tibble(
    Period = as.integer(periods[i]),
    Team = teams,
    Rating = team_elo_run$history[, i, "Rating"]
  )
})

# Build mapping: Date_num to Date
date_lookup <- games_df %>%
  distinct(Date_num, Date) %>%
  arrange(Date_num) %>%
  mutate(Period = row_number())

# Rebuild Elo history
team_elo_history <- map_dfr(seq_along(periods), function(i) {
  tibble(
    Period = as.integer(periods[i]),
    Team = teams,
    Rating = team_elo_run$history[, i, "Rating"]
  )
})

#  Join actual Date
team_elo_history <- team_elo_history %>%
  left_join(date_lookup, by = "Period")

# Plot for just "Jim Elsner & John Elsner"
team_name <- "Conrad Damon & Craig Anderson"

team_elo_history %>%
  filter(Team == team_name) %>%
  ggplot(aes(x = Period, y = Rating)) +
  geom_line(color = "blue", linewidth = 1) +
  labs(
    title = paste("Elo Rating Over Time:", team_name),
    x = "Game Period",
    y = "Elo Rating"
  ) +
  theme_minimal()

```

Compute & extract top 20 teams by Elo rating
```{r}
library(dplyr)

# Compute final Elo ratings for all teams
final_team_ratings <- team_elo_run$ratings %>%
  as_tibble() %>%
  rename(
    Team = Player,
    Rating = Rating,
    Games = Games,
    Wins = Win,
    Draws = Draw,
    Losses = Loss
  ) %>%
  arrange(desc(Rating))

# Get top 20
top_20_teams <- final_team_ratings %>%
  slice_head(n = 20)

# Print
top_20_teams
```

Nice table
```{r}
library(gt)

top_20_teams %>%
  select(Team, Rating, Games, Wins, Losses) %>%
  gt() %>%
  tab_header(
    title = "Top 20 Teams by Elo Rating"
  ) %>%
  fmt_number(columns = Rating, decimals = 0)

```
