---
title: "TallyRallyGames"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(readr)
df <- read_csv(here::here("data", "TallyRallyGames.csv"))
```

```{r}
library(dplyr)
library(stringr)
library(tidyr)

df_parsed <- df %>%
  mutate(
    # Extract temperatures
    Temp_Range = str_extract(Weather, "\\d+-\\d+\\s*F"),
    Temp_Min = as.numeric(str_extract(Temp_Range, "^\\d+")),
    Temp_Max = as.numeric(str_extract(Temp_Range, "(?<=-)\\d+")),
    Temp_Single = as.numeric(str_extract(Weather, "(?<!-)\\b\\d+(?=\\s*F)")),

    # If single temperature, use for both min & max
    Temp_Min = coalesce(Temp_Min, Temp_Single),
    Temp_Max = coalesce(Temp_Max, Temp_Single),
    Temp_Avg = rowMeans(cbind(Temp_Min, Temp_Max), na.rm = TRUE),

    # Extract wind direction (V is variable)
    Wind_Dir = str_extract(Weather, "(?i)\\b(V|SW|SE|NW|NE|S|N|E|W|WNW|ENE|NNE|SSW)\\b"),
    Wind_Dir = case_when(
      Wind_Dir == "V" ~ "Variable",
      TRUE ~ Wind_Dir
    ),

    # Extract main wind speed (range or single)
    Wind_Range = str_extract(Weather, "\\d+-\\d+\\s*mph"),
    Wind_Speed = as.numeric(str_extract(Wind_Range, "^\\d+")),
    Wind_Max1 = as.numeric(str_extract(Wind_Range, "(?<=-)\\d+")),
    Wind_Single = as.numeric(str_extract(Weather, "(?<!-)\\b\\d+(?=\\s*mph)")),

    Wind_Speed = coalesce(Wind_Speed, Wind_Single),
    Wind_Max = coalesce(Wind_Max1, Wind_Single),

    # Extract Gust if it exists
    Wind_Gust = as.numeric(str_extract(Weather, "(?<=G)\\d+")),

    # Use gust as higher wind if present
    Wind_Max = pmax(Wind_Max, Wind_Gust, na.rm = TRUE),
    Wind_Avg = rowMeans(cbind(Wind_Speed, Wind_Max), na.rm = TRUE),

    # Extract sky condition
    Sky = str_extract(Weather, "(?i)(Sunny|Partly cloudy|Cloudy|Drizzle)"),
    Sky = str_to_title(Sky)
  ) %>%
  select(
    ID, Year, `Month Day`, Round,
    `Team A Player 1`, `Team A Player 2`,
    `Team B Player 1`, `Team B Player 2`,
    `Score A`, `Score B`,
    Location, Weather,
    Temp_Min, Temp_Max, Temp_Avg,
    Wind_Dir, Wind_Speed, Wind_Gust, Wind_Max, Wind_Avg,
    Sky
  )

```

```{r}
library(dplyr)
library(tidyr)

# Ensure your scores are numeric
df_parsed <- df_parsed %>%
  mutate(
    `Score A` = as.numeric(`Score A`),
    `Score B` = as.numeric(`Score B`)
  )

# Pivot to long format for players
player_games <- df_parsed %>%
  pivot_longer(
    cols = c(`Team A Player 1`, `Team A Player 2`, `Team B Player 1`, `Team B Player 2`),
    names_to = "Player_Position",
    values_to = "Player"
  ) %>%
  mutate(
    Team = case_when(
      Player_Position %in% c("Team A Player 1", "Team A Player 2") ~ "A",
      Player_Position %in% c("Team B Player 1", "Team B Player 2") ~ "B"
    ),
    Won = case_when(
      Team == "A" & `Score A` > `Score B` ~ 1,
      Team == "B" & `Score B` > `Score A` ~ 1,
      TRUE ~ 0
    ),
    Loss = case_when(
      Team == "A" & `Score A` < `Score B` ~ 1,
      Team == "B" & `Score B` < `Score A` ~ 1,
      TRUE ~ 0
    )
  )

# Add point differential for each player per game
player_games <- player_games %>%
  mutate(
    Player_Point_Diff = case_when(
      Team == "A" ~ `Score A` - `Score B`,
      Team == "B" ~ `Score B` - `Score A`
    )
  )

# Summarize with average point differential
player_summary <- player_games %>%
  group_by(Player) %>%
  summarise(
    Games = n(),
    Wins = sum(Won, na.rm = TRUE),
    Losses = sum(Loss, na.rm = TRUE),
    Win_Pct = round(Wins / Games * 100, 1),
    Avg_Point_Diff = round(mean(Player_Point_Diff, na.rm = TRUE), 2)
  ) %>%
  arrange(desc(Win_Pct))

player_summary
```

Windy vs non-windy

```{r}
library(dplyr)
library(tidyr)

# First: classify each game as windy or not
df_parsed <- df_parsed %>%
  mutate(Windy = ifelse(Wind_Avg > 20, "Windy", "Calm"))

# Pivot into long format
player_games <- df_parsed %>%
  pivot_longer(
    cols = c(`Team A Player 1`, `Team A Player 2`, `Team B Player 1`, `Team B Player 2`),
    names_to = "Player_Position",
    values_to = "Player"
  ) %>%
  mutate(
    Team = case_when(
      Player_Position %in% c("Team A Player 1", "Team A Player 2") ~ "A",
      Player_Position %in% c("Team B Player 1", "Team B Player 2") ~ "B"
    ),
    Won = case_when(
      Team == "A" & `Score A` > `Score B` ~ 1,
      Team == "B" & `Score B` > `Score A` ~ 1,
      TRUE ~ 0
    )
  )

# Get total games played
player_totals <- player_games %>%
  group_by(Player) %>%
  summarise(Total_Games = n())

# Now summarize by Player and Windy/Calm
player_wind_summary <- player_games %>%
  group_by(Player, Windy) %>%
  summarise(
    Games = n(),
    Wins = sum(Won, na.rm = TRUE),
    Win_Pct = round(100 * Wins / Games, 1)
  ) %>%
  left_join(player_totals, by = "Player") %>%
  filter(Total_Games >= 50) %>%
  select(Player, Total_Games, Windy, Games, Wins, Win_Pct) %>%
  arrange(Player, desc(Windy))  # optional: sorts so Windy first for each player

player_wind_summary

```

Win percentage per day

```{r}
library(dplyr)

daily_summary <- player_games %>%
  filter(Player == "Jim Elsner") %>%
  group_by(`Month Day`, Wind_Avg, Temp_Avg) %>%
  summarise(
    Games = n(),
    Wins = sum(Won),
    Win_Percent = Wins / Games
  ) %>%
  arrange(`Month Day`)

daily_summary

library(ggplot2)
ggplot(daily_summary, aes(x = Wind_Avg, y = Win_Percent)) +
  geom_point(size=3) +
  geom_smooth(method="lm", se=FALSE) +
  labs(title = "Daily Win % vs Wind Speed",
       x = "Average Wind Speed (mph)",
       y = "Win Percentage")

# Linear model
lm_win <- lm(Win_Percent ~ Wind_Avg + Temp_Avg, data = daily_summary)

# Summary output
summary(lm_win)

```

```{r}
library(ggplot2)

ggplot(daily_summary, aes(x = Wind_Avg, y = Temp_Avg, z = Win_Percent)) +
  geom_contour_filled(bins = 10) +
  geom_point(aes(color = Win_Percent), size = 3) +
  scale_fill_viridis_d(option = "plasma") +
  labs(
    title = "Jim Elsner's Win Percentage by Wind & Temperature",
    x = "Average Wind Speed (mph)",
    y = "Average Temperature (F)",
    fill = "Win %"
  ) +
  theme_minimal()

```

Prediction grid

```{r}
# Get reasonable ranges from your actual data
wind_range <- seq(min(daily_summary$Wind_Avg), max(daily_summary$Wind_Avg), length.out = 50)
temp_range <- seq(min(daily_summary$Temp_Avg), max(daily_summary$Temp_Avg), length.out = 50)

# Build grid
grid <- expand.grid(
  Wind_Avg = wind_range,
  Temp_Avg = temp_range
)

# Predict win percentage
grid$Win_Percent_Pred <- predict(lm_win, newdata = grid)

library(ggplot2)

ggplot(grid, aes(x = Wind_Avg, y = Temp_Avg, z = Win_Percent_Pred)) +
  geom_contour_filled(bins = 10) +
  scale_fill_viridis_d(option = "plasma") +
  labs(
    title = "Predicted Win % for Jim Elsner",
    x = "Average Wind Speed (mph)",
    y = "Average Temperature (F)",
    fill = "Pred Win %"
  ) +
  theme_minimal()
```

A head-to-head win count matrix, showing how many times each player has defeated each other player

```{r}
library(dplyr)
library(tidyr)

# Make long results with winner and loser columns
game_results <- df_parsed %>%
  mutate(
    Winner1 = ifelse(`Score A` > `Score B` , `Team A Player 1`, `Team B Player 1`),
    Winner2 = ifelse(`Score A` > `Score B`, `Team A Player 2`, `Team B Player 2`),
    Loser1  = ifelse(`Score A` < `Score B`, `Team A Player 1`, `Team B Player 1`),
    Loser2  = ifelse(`Score A` < `Score B`, `Team A Player 2`, `Team B Player 2`)
  ) %>%
  select(Winner1, Winner2, Loser1, Loser2)

head_to_head_long <- game_results %>%
  pivot_longer(cols = c(Winner1, Winner2), names_to = "WinnerPos", values_to = "Winner") %>%
  pivot_longer(cols = c(Loser1, Loser2), names_to = "LoserPos", values_to = "Loser") %>%
  filter(Winner != Loser)  # remove self-matches

head_to_head_counts <- head_to_head_long %>%
  group_by(Winner, Loser) %>%
  summarise(Wins = n(), .groups = "drop")

head_to_head_matrix <- head_to_head_counts %>%
  pivot_wider(names_from = Loser, values_from = Wins, values_fill = 0)
```

Elo ratings: Why is it powerful? It automatically adjusts for strength of opponent: beating top players moves you up fast, beating weak players not much. Gives you a single number to rank players, but it implicitly accounts for all games played. Great for dynamic tracking: you can plot Elo vs time to see if someone’s getting better or declining.

```{r}
library(readr)
games_df <- read_csv(here::here("data", "TallyRallyGames.csv"))

library(dplyr)
library(tidyr)
library(PlayerRatings)

# Create all winner vs loser records
pairwise_results <- games_df %>%
  filter(!is.na(`Score A`) & !is.na(`Score B`)) %>%
  mutate(
    Winning_P1 = if_else(`Score A` > `Score B`, `Team A Player 1`, `Team B Player 1`),
    Winning_P2 = if_else(`Score A` > `Score B`, `Team A Player 2`, `Team B Player 2`),
    Losing_P1  = if_else(`Score A` < `Score B`, `Team A Player 1`, `Team B Player 1`),
    Losing_P2  = if_else(`Score A` < `Score B`, `Team A Player 2`, `Team B Player 2`)
  ) %>%
  select(Winning_P1, Winning_P2, Losing_P1, Losing_P2) %>%
  pivot_longer(cols = c(Winning_P1, Winning_P2), names_to = "win_role", values_to = "Winner") %>%
  pivot_longer(cols = c(Losing_P1, Losing_P2), names_to = "lose_role", values_to = "Loser") %>%
  transmute(Date = 1, Player1 = Winner, Player2 = Loser, Score = 1) %>%
  distinct()

# Now add inverse games (losses count as 0)
elo_input <- bind_rows(
  pairwise_results,
  pairwise_results %>%
    mutate(Player1 = Player2, Player2 = Player1, Score = 0)
)

# Run Elo model
elo_ratings <- elo(elo_input, init = 1500, k = 32)

# Show top 25 players
elo_ratings$ratings %>%
  arrange(desc(Rating)) %>%
  mutate(Rating = round(Rating, 0)) %>%
  slice_head(n = 25)
```

The **Elo rating system** is a way to estimate the skill levels of players in competitive games.

It was invented by **Arpad Elo**, a Hungarian-American physics professor and chess master.\
The system became widely used in chess, but is now used for many sports and games, including online gaming and even evaluating football teams.

The system works like this:

-   Each player starts with an initial rating (commonly 1500).
-   When two players (or teams) play a match:
    -   The expected outcome is calculated based on the difference in their ratings.
    -   If a player performs better than expected (wins against a stronger opponent), they gain more points.
    -   If a player performs worse than expected (loses to a weaker opponent), they lose more points.
-   The **K-factor** controls how much the ratings adjust after each game.

The expected score (probability of winning) is:

$$
E_A = \frac{1}{1 + 10^{(R_B - R_A) / 400}}
$$

where: - $R_A$ is Player A’s rating - $R_B$ is Player B’s rating

Then after the match:

$$
R'_A = R_A + K \times (S_A - E_A)
$$

where: $S_A = 1$ if A wins, $0.5$ if draw, $0$ if A loses and $K$ is the adjustment factor (commonly 32).

✅ It’s **self-correcting**:\
If a player wins consistently, their rating rises, so future wins give less of a boost (and losses hurt more).

✅ It’s **continuous**:\
It tracks performance game by game, without needing to wait for a whole season.

✅ It’s **predictive**:\
A higher rating difference means a higher expected probability of winning.

-   Players with the same rating are expected to win about equally often.
-   The ratings adjust automatically to how surprising the outcomes are.
-   It can handle long histories of matchups without starting over.

You’re using this system to track and compare players like **Jim Elsner**, **Mike Prost**, and others across multiple tournaments.

By plotting Elo over time, you can see:

-   who’s improving
-   who’s consistent
-   who might be having up or down seasons.

## References

-   Arpad Elo, *The Rating of Chessplayers, Past and Present*, 1978.
-   FIDE Handbook on Elo calculation: <https://handbook.fide.com/chapter/B022017>
-   Wikipedia: <https://en.wikipedia.org/wiki/Elo_rating_system>

Elo Escape!! ratings over time

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(PlayerRatings)
library(here)

# Read your data
games_df <- read_csv(here("data", "TallyRallyGames.csv"))

# Create proper Date and numeric time index
games_df <- games_df %>%
  mutate(
    Month_Day_Clean = str_remove(`Month Day`, "(st|nd|rd|th)"),
    Date_str = paste(Month_Day_Clean, Year),
    Date = mdy(Date_str)
  )

# Create numeric index (e.g. days since first date)
games_df <- games_df %>%
  arrange(Date) %>%
  mutate(Date_num = as.integer(difftime(Date, min(Date, na.rm = TRUE), units = "days")))

# Build pairwise winner-loser data
pairwise_results <- games_df %>%
  filter(!is.na(`Score A`) & !is.na(`Score B`)) %>%
  mutate(
    Winning_P1 = if_else(`Score A` > `Score B`, `Team A Player 1`, `Team B Player 1`),
    Winning_P2 = if_else(`Score A` > `Score B`, `Team A Player 2`, `Team B Player 2`),
    Losing_P1  = if_else(`Score A` < `Score B`, `Team A Player 1`, `Team B Player 1`),
    Losing_P2  = if_else(`Score A` < `Score B`, `Team A Player 2`, `Team B Player 2`)
  ) %>%
  select(Date_num, Winning_P1, Winning_P2, Losing_P1, Losing_P2) %>%
  pivot_longer(c(Winning_P1, Winning_P2), names_to = "win_role", values_to = "Winner") %>%
  pivot_longer(c(Losing_P1, Losing_P2), names_to = "lose_role", values_to = "Loser") %>%
  transmute(Date = Date_num, Player1 = Winner, Player2 = Loser, Score = 1) %>%
  distinct()

# Build input for Elo (losses as 0)
elo_input <- bind_rows(
  pairwise_results,
  pairwise_results %>%
    mutate(Player1 = Player2, Player2 = Player1, Score = 0)
)

# Run Elo model
elo_run <- elo(elo_input, init = 1500, k = 32, history = TRUE)

#  Print final ratings
elo_ratings$ratings %>%
  arrange(desc(Rating)) %>%
  mutate(Rating = round(Rating, 0)) %>%
  slice_head(n = 25)
```

```{r}
library(purrr)

# Extract dim names
players <- dimnames(elo_run$history)[[1]]
periods <- dimnames(elo_run$history)[[2]]

# Build tidy data frame
elo_history <- map_dfr(seq_along(periods), function(i) {
  tibble(
    Period = as.integer(periods[i]),
    Player = players,
    Rating = elo_run$history[, i, "Rating"]
  )
})

# Filter to your players
players_to_plot <- c("Jim Elsner", "Diana Elsner", "JaSun Burdick", "Craig Anderson", "Hank Bass", "Drew Magee", "Jordan Huston", "Scott Schmucker", "Conrad Damon", "Harvey Brandt", "John Elsner", "Mike Prost")
elo_history_filtered <- elo_history %>%
  filter(Player %in% players_to_plot)

library(lubridate)

# Reconstruct dates from Date_num
period_dates <- elo_input %>%
  distinct(Date) %>%
  arrange(Date) %>%
  mutate(Period = row_number(),
         Actual_Date = min(games_df$Date, na.rm = TRUE) + days(Date)) %>%
  select(Period, Actual_Date)

elo_history_with_date <- elo_history %>%
  left_join(period_dates, by = "Period")

ggplot(elo_history_with_date %>% filter(Player %in% players_to_plot),
       aes(x = Actual_Date, y = Rating, color = Player)) +
  geom_line(alpha = 0.7) +  # main lines
  geom_smooth(se = FALSE, method = "loess", span = 0.75) +  # smoothed trend line
  facet_wrap(~ Player, scales = "fixed", ncol = 3) +  # small multiples
  labs(
    title = "Elo Ratings Over Time by Player",
    x = "Date",
    y = "Elo Rating"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
