---
title: "Player team network"
output: html_document
date: "2025-05-12"
editor_options: 
  chunk_output_type: console
---

Build a network graph of Double Disc Court (DDC) players and their teammates to visualize connections in the competitive community

Working with data from https://doubledisccourt.com/results, the basic idea is to: Scrape the event pages to extract teams and players, count participation frequency to identify the top 25 players, build a graph where each edge represents a team, visualize the network in R using {igraph} and optionally {ggraph}

Install and load required libraries
```{r}
install.packages(c("rvest", "dplyr", "stringr", "igraph", "ggraph", "tidygraph"))

library(rvest)
library(dplyr)
library(stringr)
library(igraph)
library(ggraph)
library(tidygraph)
```

Scrape team data from each event. Start with one tournament url

```{r}
library(chromote)
library(rvest)

# Load tournament page
url <- "https://doubledisccourt.com/results/tournament.html?id=603"
b <- ChromoteSession$new()
b$Page$navigate(url)
Sys.sleep(5)  # wait for JS to populate the page

# Extract rendered HTML
html_doc <- b$DOM$getDocument()
source <- b$DOM$getOuterHTML(nodeId = html_doc$root$nodeId)[["outerHTML"]]
page <- read_html(source)

# Extract <li> elements in the tournament result list
teams <- page %>% html_nodes("ul.form li")

# Skip the header row and parse player names
team_data <- lapply(teams[-1], function(team) {
  cells <- team %>% html_nodes("div.dataCell")
  if (length(cells) >= 2) {
    names <- cells[2] %>%
      html_nodes("a") %>%
      html_text(trim = TRUE)
    
    if (length(names) == 2) {
      data.frame(Player1 = names[1], Player2 = names[2], stringsAsFactors = FALSE)
    } else {
      NULL
    }
  } else {
    NULL
  }
})

# Combine into a data frame
df <- bind_rows(team_data)
print(df)
```

Loop over all event pages
```{r}
library(purrr)

# Function to parse one event
scrape_ddc_event <- function(id) {
  print(id)
  url <- paste0("https://doubledisccourt.com/results/tournament.html?id=", id)
  b <- ChromoteSession$new()
  b$Page$navigate(url)
  Sys.sleep(5)

  html_doc <- b$DOM$getDocument()
  source <- b$DOM$getOuterHTML(nodeId = html_doc$root$nodeId)[["outerHTML"]]
  b$close()

  page <- read_html(source)

  # Check for King-style or solo-format events
  if (grepl("King of the Court|Monarch of the Court|Solo format", page %>% html_text(), ignore.case = TRUE)) {
    return(NULL)
  }

  # Extract year from <title>
  title_text <- page %>% html_node("title") %>% html_text()
  year_match <- str_extract(title_text, "\\b(19|20)\\d{2}\\b")
  event_year <- ifelse(is.na(year_match), NA, as.integer(year_match))

  # Get all division headers and corresponding result lists
  headers <- page %>% html_nodes("div.header")
  result_lists <- page %>% html_nodes("ul.form")

  # Safety check
  if (length(headers) == 0 || length(headers) != length(result_lists)) return(NULL)

  all_divisions <- list()

  for (i in seq_along(headers)) {
    div_name <- headers[i] %>% html_text(trim = TRUE)
    teams <- result_lists[[i]] %>% html_nodes("li")

    if (length(teams) < 2) next

    # Parse each <li> in the division's result list
    team_data <- lapply(teams[-1], function(team) {
      cells <- team %>% html_nodes("div.dataCell")
      if (length(cells) >= 2) {
        place <- cells[1] %>% html_text(trim = TRUE)
        names <- cells[2] %>% html_nodes("a") %>% html_text(trim = TRUE)

        if (length(names) == 2) {
          return(data.frame(
            Year = event_year,
            Division = div_name,
            Place = place,
            Player1 = names[1],
            Player2 = names[2],
            TournamentID = id,
            stringsAsFactors = FALSE
          ))
        }
      }
      return(NULL)
    })

    all_divisions[[i]] <- bind_rows(team_data)
  }

  df <- bind_rows(all_divisions)
  if (nrow(df) == 0) return(NULL)
  return(df)
}

# Loop through a range of event IDs
event_ids <- 1:602

all_teams <- map_dfr(event_ids, possibly(scrape_ddc_event, NULL))

# Save to CSV
write.csv(all_teams, "data/output/ddc_team_data.csv", row.names = FALSE)

# View a sample
print(head(all_teams))

```

Code to create a plot the network graph
```{r}
library(dplyr)
library(igraph)
library(ggraph)
library(tidygraph)

# Step 1: Filter to Open Division since xxxx
filtered_data <- all_teams %>%
  filter(Division == "Open", Year >= 1995)

# Step 2: Count pairings (treat A-B and B-A as the same team)
pair_counts <- filtered_data %>%
  rowwise() %>%
  mutate(
    player_a = min(Player1, Player2),
    player_b = max(Player1, Player2)
  ) %>%
  ungroup() %>%
  count(player_a, player_b, name = "team_count")

# Step 2.5: filter counts
pair_counts_strong <- pair_counts %>%
  filter(team_count >= 5)

# Step 3: Build igraph object
g <- graph_from_data_frame(pair_counts_strong, directed = FALSE)

# Step 4: Basic visualization with igraph
plot(g,
     vertex.size = 5,
     edge.width = E(g)$team_count,
     vertex.label.cex = 0.8,
     main = "DDC Open Division Player Network (Since 2010)")

# Optional: fancier layout with ggraph
tg <- as_tbl_graph(g)

ggraph(tg, layout = "fr") +
  geom_edge_link(aes(width = team_count), alpha = 0.3) +
  geom_node_point(color = "tomato", size = 4) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_edge_width_continuous(
    breaks = c(5, 10, 20, 40),
    labels = c("5", "10", "20", "40"),
    name = "Count"
  ) +
  labs(title = " DDC Open Division Player Network (1995–present, 5+ Pairings)") +
  theme_void()
```

By decade
```{r}
library(dplyr)
library(igraph)
library(ggraph)
library(tidygraph)
library(RColorBrewer)

# Define decade ranges and output folder
decades <- list(
  "1985-1994" = 1985:1994,
  "1995–2004" = 1995:2004,
  "2005–2014" = 2005:2014,
  "2015–2024" = 2015:2024
)

# Create an output folder
output_dir <- "data/output/ddc_decade_networks"
dir.create(output_dir, showWarnings = FALSE)

# Loop through decades and generate plots
for (label in names(decades)) {
  years <- decades[[label]]

  df <- all_teams %>%
    filter(Division == "Open", Year %in% years) %>%
    rowwise() %>%
    mutate(
      player_a = min(Player1, Player2),
      player_b = max(Player1, Player2)
    ) %>%
    ungroup() %>%
    count(player_a, player_b, name = "team_count") %>%
    filter(team_count >= 3)

  if (nrow(df) == 0) {
    message(paste("No data for", label, "- skipping"))
    next
  }

  # Build graph
  g <- graph_from_data_frame(df, directed = FALSE)
  tg <- as_tbl_graph(g)
  E(tg)$team_count <- df$team_count

  # Save to PNG
  file_name <- paste0(output_dir, "/ddc_network_", gsub("–", "-", label), ".png")
  png(file_name, width = 1200, height = 900, res = 150)

print(
  ggraph(tg, layout = "fr") +
    geom_edge_link(aes(width = team_count), color = "gray", alpha = 0.35, show.legend = FALSE) +
    geom_node_point(color = "tomato", size = 3) +
    geom_node_text(aes(label = name), repel = TRUE, size = 2.8) +
    scale_edge_width_continuous(
      breaks = c(5, 10, 20, 40),
      labels = c("5", "10", "20", "40"),
      name = "Count"
    ) +
#    labs(title = paste(" DDC Open Division Partner Network (3+ Pairings)", label)) +
    theme_void()
)

  dev.off()
  message("Saved ", file_name)
}

```

Combine images
```{r}
library(magick)

# Helper: read, add border, add title
process_image <- function(path, title, border_color = "gray", border_size = "5x5") {
  image_read(path) %>%
    image_border(border_color, border_size) %>%
    image_annotate(
      title,
      size = 40,
      location = "+15+15",
      gravity = "northwest",
      color = "gray",
      font = "Arial",
      weight = 700
    )
}

# Step 1: Load and process each image
img1 <- process_image("data/output/ddc_decade_networks/ddc_network_1985-1994.png", "1985–1994")
img2 <- process_image("data/output/ddc_decade_networks/ddc_network_1995-2004.png", "1995–2004")
img3 <- process_image("data/output/ddc_decade_networks/ddc_network_2005-2014.png", "2005–2014")
img4 <- process_image("data/output/ddc_decade_networks/ddc_network_2015-2024.png", "2015–2024")

# Step 2: Append into 2 rows
row1 <- image_append(c(img1, img2))
row2 <- image_append(c(img3, img4))

# Step 3: Stack vertically for 2 × 2 grid
final_image <- image_append(c(row1, row2), stack = TRUE)

# Step 4: Create a title banner image
title_banner <- image_blank(
  width = image_info(final_image)$width,
  height = 100,
  color = "white"
) %>%
  image_annotate(
    "DDC Open Division Player Networks by Decade (3+ Pairings)",
    size = 50,
    gravity = "center",
    color = "black",
    font = "Arial",
    weight = 700
  )

# Step 5: Stack title + grid
full_layout <- image_append(c(title_banner, final_image), stack = TRUE)

# Step 6: Save final image
image_write(full_layout, path = "data/output/ddc_decade_networks/ddc_networks_2x2_full_title.png", format = "png")
```


```{r}
install.packages("magick")
library(magick)

img1 <- image_read("data/output/ddc_decade_networks/ddc_network_1985-1994.png")
img2 <- image_read("data/output/ddc_decade_networks/ddc_network_1995-2004.png")
img3 <- image_read("data/output/ddc_decade_networks/ddc_network_2005-2014.png")
img4 <- image_read("data/output/ddc_decade_networks/ddc_network_2015-2024.png")

# First row: 1985–1994 and 1995–2004
row1 <- image_append(c(img1, img2))

# Second row: 2005–2014 and 2015–2024
row2 <- image_append(c(img3, img4))

# Stack rows vertically
final_image <- image_append(c(row1, row2), stack = TRUE)
image_write(final_image, path = "data/output/ddc_decade_networks/ddc_network_2x2.png", format = "png")
```

Ego-centric network
```{r}
library(dplyr)
library(igraph)
library(ggraph)
library(tidygraph)

# Set your target player
target_player <- "Jim Elsner"

# Filter all pairings involving the player
ego_pairs <- all_teams %>%
  filter(Division == "Open") %>%
  filter(Player1 == target_player | Player2 == target_player) %>%
  mutate(
    teammate = ifelse(Player1 == target_player, Player2, Player1)
  ) %>%
  count(teammate, name = "team_count")

# Build edge list (target → teammate)
edges <- data.frame(
  from = target_player,
  to = ego_pairs$teammate,
  weight = ego_pairs$team_count
)

# Build graph
g <- graph_from_data_frame(edges, directed = FALSE)
tg <- as_tbl_graph(g)

# Plot
ggraph(tg, layout = "star", center = target_player) +
  geom_edge_link(aes(width = weight), color = "gray30", alpha = 0.7) +
  geom_node_point(size = 5, color = ifelse(V(g)$name == target_player, "tomato", "steelblue")) +
  geom_node_text(aes(label = name), repel = TRUE, size = 4) +
  scale_edge_width_continuous(
    breaks = c(1, 3, 5, 10),
    labels = c("1", "3", "5", "10"),
    name = "Teamings"
  ) +
  labs(title = paste("Ego Network for", target_player)) +
  theme_void()
```

Two-layer ego-centric network
```{r}
library(dplyr)
library(igraph)
library(ggraph)
library(tidygraph)

# Ego
target_player <- "Diana Elsner"

# Base pairings (deduplicated)
ddc_pairs <- all_teams %>%
#  filter(Division == "Open") %>%
  mutate(
    player_a = pmin(Player1, Player2),
    player_b = pmax(Player1, Player2)
  ) %>%
  distinct(player_a, player_b)

# Step 1: Ego's direct teammates
ego_edges <- ddc_pairs %>%
  filter(player_a == target_player | player_b == target_player) %>%
  mutate(teammate = ifelse(player_a == target_player, player_b, player_a))

teammates <- unique(ego_edges$teammate)

# Step 2: All pairings where BOTH players are in teammates list (i.e. teammate-to-teammate or teammate-to-others)
second_level_edges <- ddc_pairs %>%
  filter(player_a %in% teammates | player_b %in% teammates) %>%
  filter(player_a != target_player & player_b != target_player)

# Step 3: Combine ego links and second-degree links
all_edges <- bind_rows(
  ego_edges %>% select(from = player_a, to = player_b),
  second_level_edges %>% rename(from = player_a, to = player_b)
) %>%
  filter(!is.na(from), !is.na(to))

# Step 4: Build and plot
g <- graph_from_data_frame(all_edges, directed = FALSE)
tg <- as_tbl_graph(g)

ggraph(tg, layout = "fr") +
  geom_edge_link(color = "gray70", alpha = 0.6) +
  geom_node_point(aes(color = name == target_player), size = 5) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_color_manual(values = c("steelblue", "tomato"), guide = "none") +
  labs(title = paste("2-Level Ego Network for", target_player)) +
  theme_void()

```

Partnership “reach” hierarchy for each player
```{r}
library(dplyr)
library(igraph)
library(tidygraph)

# STEP 1: Build full player network
pair_df <- all_teams %>%
  filter(Division == "Open") %>%
  mutate(
    player_a = pmin(Player1, Player2),
    player_b = pmax(Player1, Player2)
  ) %>%
  distinct(player_a, player_b)

# Create undirected graph
g <- graph_from_data_frame(pair_df, directed = FALSE)
tg <- as_tbl_graph(g)

# STEP 2: For each player, compute 1st, 2nd, 3rd-level neighbors
all_players <- sort(unique(c(pair_df$player_a, pair_df$player_b)))

# Function to compute levels
get_partner_levels <- function(player_name) {
  ego_id <- which(V(g)$name == player_name)
  if (length(ego_id) == 0) return(NULL)

  # Level 1 (direct teammates)
  n1 <- ego(g, order = 1, nodes = ego_id, mode = "all")[[1]] %>% as_ids()
  
  # Level 2 (partners of direct teammates)
  n2 <- ego(g, order = 2, nodes = ego_id, mode = "all")[[1]] %>% as_ids()
  n2 <- setdiff(n2, c(player_name, n1))

  # Level 3 (partners of 2nd-level players)
  n3 <- ego(g, order = 3, nodes = ego_id, mode = "all")[[1]] %>% as_ids()
  n3 <- setdiff(n3, c(player_name, n1, n2))

  data.frame(
    Player = player_name,
    FirstLevel = length(n1),
    SecondLevel = length(n2),
    ThirdLevel = length(n3)
  )
}

# STEP 3: Apply to all players
partner_levels_df <- do.call(rbind, lapply(all_players, get_partner_levels))

# STEP 4: Sort by number of first-level partners
partner_levels_df <- partner_levels_df %>%
  arrange(desc(FirstLevel))

# View the result
head(partner_levels_df)
```

Make the ego graph a Shiny app
```{r}
library(shiny)
library(igraph)
library(ggraph)
library(tidygraph)
library(dplyr)

# Load DDC data
ddc_data <- read.csv("data/output/ddc_team_data.csv", stringsAsFactors = FALSE)

# UI
ui <- fluidPage(
  titlePanel("Double Disc Court Network (ego) Viewer"),

  sidebarLayout(
    sidebarPanel(
      selectInput("division", "Division:", choices = c("Any", "Open", "Women"), selected = "Open"),
      sliderInput("year_range", "Year range:",
                  min = min(ddc_data$Year, na.rm = TRUE),
                  max = max(ddc_data$Year, na.rm = TRUE),
                  value = c(2010, 2024), sep = "", step = 1),
      selectInput("player", "Select a player:", choices = NULL),
      numericInput("level", "Network depth (1 = teammates only, 2 = teammates-of-teammates):",
                   value = 2, min = 1, max = 3)
    ),

    mainPanel(
      plotOutput("egoPlot", height = "700px")
    )
  )
)

# SERVER
server <- function(input, output, session) {

  # Reactive: Filter data by division and year
  filtered_data <- reactive({
    df <- ddc_data %>%
      filter(Year >= input$year_range[1], Year <= input$year_range[2])

    if (input$division != "Any") {
      df <- df %>% filter(Division == input$division)
    }

    df
  })

  # Reactive: update player list based on filtered data
  observe({
  players <- sort(unique(c(filtered_data()$Player1, filtered_data()$Player2)))
  current <- input$player
  if (!is.null(current) && current %in% players) {
    selected <- current
  } else {
    selected <- players[1]
  }
  updateSelectInput(session, "player", choices = players, selected = selected)
})

  # Reactive: Build graph
  graph_data <- reactive({
    df <- filtered_data()

    df %>%
      mutate(
        player_a = pmin(Player1, Player2),
        player_b = pmax(Player1, Player2)
      ) %>%
      distinct(player_a, player_b)
  })

  g <- reactive({
    graph_from_data_frame(graph_data(), directed = FALSE)
  })

  output$egoPlot <- renderPlot({
    req(input$player)
    net <- g()
    ego_id <- which(V(net)$name == input$player)

    if (length(ego_id) == 0) return(NULL)

    # Ego network subgraph
    sub_nodes <- ego(net, order = input$level, nodes = ego_id, mode = "all")[[1]]
    subg <- induced_subgraph(net, vids = sub_nodes)
    tg <- as_tbl_graph(subg)

    # Plot
    ggraph(tg, layout = "fr") +
      geom_edge_link(color = "gray70", alpha = 0.6) +
      geom_node_point(aes(color = name == input$player), size = 5) +
      geom_node_text(aes(label = name), repel = TRUE, size = 3) +
      scale_color_manual(values = c("steelblue", "tomato"), guide = "none") +
      labs(title = paste("Ego Network for", input$player)) +
      theme_void()
  })
}

# Run the app
shinyApp(ui = ui, server = server)
```

Once this successfully runs locally then try to deploy it. Copy the entire code chunk to a file called app.R. Put that file in a separate directory. Put your data in that same directory. Change the read.csv() function to point to this new directory.

Deploy it
```{r}
install.packages("rsconnect")

library(rsconnect)

rsconnect::setAccountInfo(name='jameselsner',
			  token='C715F8CF8311C0BE753AC0545FA37D56',
			  secret='gkgAYlb5SCVKZDIaQb4VG3Vr/23xQOXp8urmMaRj')



```



