---
title: "Player Debut Timeline"
output: html_document
date: "2025-07-03"
editor_options: 
  chunk_output_type: console
---

Given  `all_tournaments_df`, here’s what we’ll do:

1. Reshape the data from wide to long so Player1 and Player2 are combined in a single Player column.
2. Filter out any NA players (important for solo tournaments that might only list Player1).
3. Keep only data from 1977 on.
4. Find the first year each player appeared.
5. Sort by year and player name, and assign an ordinal count.

```{r}
all_tournaments_df <- read_csv(here::here("data", "output", "ddc_full_tournament_data_auto.csv"))
```

```{r}
library(dplyr)
library(tidyr)

# Build player debut data
player_debuts <- all_tournaments_df %>%
#  filter(Location == "Tallahassee") %>%
  filter(Year >= 1977) %>%
  select(EventStartDate, Player1, Player2, Location) %>%
  pivot_longer(cols = c(Player1, Player2), names_to = "Role", values_to = "Player") %>%
  filter(!is.na(Player)) %>%
  group_by(Player) %>%
  summarise(FirstDate = min(EventStartDate),
            Location = first(Location)) %>%
  arrange(FirstDate, Player) %>%
  mutate(Count = row_number())

# Now expand this out into a timeline by player debut year
timeline_df <- player_debuts %>%
  select(Date = FirstDate, Player, Count, Location)

# View the result
print(timeline_df)
```

Plots
```{r}
library(ggplot2)

ggplot(timeline_df, aes(x = Date, y = Count)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "tomato", size = 2) +
  labs(
    title = "Cumulative Number of New DDC Tournament Players Over Time",
    x = "Year",
    y = "Cumulative Players",
    caption = "Data from https://doubledisccourt.com/results/rankings.html"
  ) +
  theme_minimal()
```

Number of events per player
```{r}
library(dplyr)
library(tidyr)
library(stringr)

player_event_counts <- all_tournaments_df %>%
  filter(!str_detect(tolower(TournamentName), "league")) %>%  # remove "league"
  select(TournamentID, Player1, Player2) %>%
  pivot_longer(cols = c(Player1, Player2), names_to = "Role", values_to = "Player") %>%
  filter(!is.na(Player)) %>%
  distinct(TournamentID, Player) %>%
  count(Player, name = "NumEvents") %>%
  arrange(desc(NumEvents))

# Show results
print(player_event_counts)

```

Number of events per team
```{r}
library(dplyr)
library(stringr)

team_counts <- all_tournaments_df %>%
  filter(!str_detect(tolower(TournamentName), "league")) %>%  # exclude leagues
  filter(!is.na(Player1) & !is.na(Player2)) %>%               # keep only proper teams
  mutate(
    player_a = pmin(Player1, Player2),  # order names alphabetically so (Jim, Mike) == (Mike, Jim)
    player_b = pmax(Player1, Player2)
  ) %>%
  distinct(TournamentID, player_a, player_b) %>%              # ensure counted once per tournament
  count(player_a, player_b, name = "NumEvents") %>%
  arrange(desc(NumEvents))

# View top pairs
print(team_counts)

```

Player partnership graph
```{r}
library(igraph)
library(ggraph)
library(tidygraph)

# Filter to strong partnerships
strong_teams <- team_counts %>%
  filter(NumEvents >= 8)

# Build graph
g <- graph_from_data_frame(strong_teams, directed = FALSE)

# Plot
ggraph(as_tbl_graph(g), layout = "fr") +
  geom_edge_link(aes(width = NumEvents), color = "gray60", alpha = 0.6) +
  geom_node_point(color = "tomato", size = 4) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_edge_width_continuous(range = c(0.5, 5)) +
  labs(
    title = "DDC Strong Player Partnership Network",
    subtitle = "Showing only partnerships with at least 8 tournaments together"
  ) +
  theme_void()

```

Color nodes by number of unique partners
```{r}
# Compute overall degree from full team_counts graph
full_graph <- graph_from_data_frame(team_counts, directed = FALSE)
full_degree <- degree(full_graph)
full_degree_df <- data.frame(
  Player = names(full_degree),
  TotalPartners = as.numeric(full_degree)
)

# Filter to strong partnerships
strong_teams <- team_counts %>%
  filter(NumEvents >= 8)

strong_graph <- graph_from_data_frame(strong_teams, directed = FALSE)

library(tidygraph)

tg_strong <- as_tbl_graph(strong_graph) %>%
  left_join(full_degree_df, by = c("name" = "Player"))

library(ggraph)

ggraph(tg_strong, layout = "fr") +
  geom_edge_link(aes(width = NumEvents), color = "gray70", alpha = 0.5) +
  geom_node_point(aes(color = TotalPartners), size = 5) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_edge_width(range = c(0.5, 5)) +
  scale_color_viridis_c(option = "plasma", name = "Total Unique Partners") +
  labs(
    title = "DDC Strong Player Partnership Network",
    subtitle = "Node color reflects total unique partners across all tournaments"
  ) +
  theme_void()

```





