---
title: "Player team network"
output: html_document
date: "2025-05-12"
editor_options: 
  chunk_output_type: console
---

Build a network graph of Double Disc Court (DDC) players and their teammates to visualize connections in the competitive community

Working with data from https://doubledisccourt.com/results, the basic idea is to: read the already-scraped team data, count partnership frequency, build a graph where each edge represents a team, and visualize the network in R using {igraph} and {ggraph}.

The scraping workflow is handled outside this notebook by the update script. Run this when the source website has new tournaments:

```sh
Rscript scripts/update_ddc_tournament_data.R
```

Load required libraries and data
```{r}
library(dplyr)
library(readr)
library(igraph)
library(ggraph)
library(tidygraph)
library(here)
```

```{r}
all_teams <- read_csv(
  here("data", "output", "ddc_team_data.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    Place = as.character(Place),
    Player1 = trimws(Player1),
    Player2 = trimws(Player2)
  ) %>%
  filter(
    !is.na(Player1),
    !is.na(Player2),
    Player1 != "",
    Player2 != ""
  )

glimpse(all_teams)
```

Code to create and plot the network graph
```{r}
library(dplyr)
library(igraph)
library(ggraph)
library(tidygraph)

# Step 1: Filter to Open Division since xxxx
filtered_data <- all_teams %>%
  filter(Division == "Open", Year >= 2021, Place %in% c("1", "2", "3"))

# Step 2: Count pairings (treat A-B and B-A as the same team)
pair_counts <- filtered_data %>%
  rowwise() %>%
  mutate(
    player_a = min(Player1, Player2),
    player_b = max(Player1, Player2)
  ) %>%
  ungroup() %>%
  count(player_a, player_b, name = "team_count")

# Step 2.5: filter counts
pair_counts_strong <- pair_counts %>%
  filter(team_count >= 1)

# Step 3: Build igraph object
g <- graph_from_data_frame(pair_counts_strong, directed = FALSE)
tg <- as_tbl_graph(g)

ggraph(tg, layout = "fr") +
#  geom_edge_link(aes(width = team_count), alpha = 0.3) +
  geom_edge_link(alpha = 0.3) +
  geom_node_point(color = "tomato", size = 4) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
#  scale_edge_width_continuous(
#    breaks = c(5, 10, 20, 40),
#    labels = c("5", "10", "20", "40"),
#    name = "Count"
#  ) +
  labs(title = " DDC Open Division Player Network (2021–present)") +
  theme_void()
```

Extract and sort node metrics
```{r}
library(igraph)
library(dplyr)

# Assume g is your undirected graph of player pairings
# Each node = player

# Compute metrics
node_df <- data.frame(
  Player = V(g)$name,
  Degree = degree(g),
  Betweenness = betweenness(g, normalized = TRUE)
)

# Sort by Degree descending
node_df_sorted <- node_df %>%
  arrange(desc(Degree))

# View the top
print(node_df_sorted)
```

What is Betweenness Centrality? Betweenness centrality measures:

How often a player lies on the shortest path between other players. In simpler terms: A player with high betweenness is a bridge connecting different parts of the network. They may not have the most teammates, but they connect teammate groups that otherwise wouldn’t be connected.

Let’s say Jim Elsner played with people in both the Wisconsin and Florida DDC scenes. Those two groups don’t team with each other often. Jim will lie on the shortest path between many Wisconsin–Florida player pairs. Result: High betweenness

Unnormalized betweenness is the raw count of how many shortest paths a node lies on. Normalized betweenness divides this count by the maximum possible number of shortest paths a node could be on — making the value:
So in the DDC network: A normalized betweenness of 0.15 means the player is on 15% of all shortest paths between other players (after adjusting for network size)

Why Normalize?
Makes betweenness scores comparable across networks/ Helpful if you're looking at decade-based subgraphs or player subsets

Why it Matters in DDC: Players with high normalized betweenness: Often serve as connectors between regions or generations. May have been mentors, organizers, or simply active in multiple scenes/ Might not have the most direct partners (degree), but they’re strategic hubs

```{r}
library(ggraph)
library(tidygraph)
library(igraph)
# Add normalized betweenness as a node attribute
tg <- tg %>%
  mutate(Betweenness = centrality_betweenness(normalized = TRUE))

# Plot
ggraph(tg, layout = "fr") +
  geom_edge_link(color = "gray70", alpha = 0.6) +
  geom_node_point(aes(size = Betweenness), color = "steelblue", alpha = 0.8) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_size_continuous(range = c(2, 12), name = "Betweenness") +
  labs(title = "DDC Player Network\nNode Size = Betweenness Centrality") +
  theme_void()
```

By decade
```{r}
library(dplyr)
library(igraph)
library(ggraph)
library(tidygraph)
library(RColorBrewer)

# Define decade ranges and output folder
decades <- list(
  "1985-1994" = 1985:1994,
  "1995–2004" = 1995:2004,
  "2005–2014" = 2005:2014,
  "2015–2024" = 2015:2024
)

# Create an output folder
output_dir <- "data/output/ddc_decade_networks"
dir.create(output_dir, showWarnings = FALSE)

# Loop through decades and generate plots
for (label in names(decades)) {
  years <- decades[[label]]

  df <- all_teams %>%
    filter(Division == "Open", Year %in% years) %>%
    rowwise() %>%
    mutate(
      player_a = min(Player1, Player2),
      player_b = max(Player1, Player2)
    ) %>%
    ungroup() %>%
    count(player_a, player_b, name = "team_count") %>%
    filter(team_count >= 3)

  if (nrow(df) == 0) {
    message(paste("No data for", label, "- skipping"))
    next
  }

  # Build graph
  g <- graph_from_data_frame(df, directed = FALSE)
  tg <- as_tbl_graph(g)
  E(tg)$team_count <- df$team_count

  # Save to PNG
  file_name <- paste0(output_dir, "/ddc_network_", gsub("–", "-", label), ".png")
  png(file_name, width = 1200, height = 900, res = 150)

print(
  ggraph(tg, layout = "fr") +
    geom_edge_link(aes(width = team_count), color = "gray", alpha = 0.35, show.legend = FALSE) +
    geom_node_point(color = "tomato", size = 3) +
    geom_node_text(aes(label = name), repel = TRUE, size = 2.8) +
    scale_edge_width_continuous(
      breaks = c(5, 10, 20, 40),
      labels = c("5", "10", "20", "40"),
      name = "Count"
    ) +
#    labs(title = paste(" DDC Open Division Partner Network (3+ Pairings)", label)) +
    theme_void()
)

  dev.off()
  message("Saved ", file_name)
}

```

Combine images
```{r}
library(magick)

# Helper: read, add border, add title
process_image <- function(path, title, border_color = "gray", border_size = "5x5") {
  image_read(path) %>%
    image_border(border_color, border_size) %>%
    image_annotate(
      title,
      size = 40,
      location = "+15+15",
      gravity = "northwest",
      color = "gray",
      font = "Arial",
      weight = 700
    )
}

# Step 1: Load and process each image
img1 <- process_image("data/output/ddc_decade_networks/ddc_network_1985-1994.png", "1985–1994")
img2 <- process_image("data/output/ddc_decade_networks/ddc_network_1995-2004.png", "1995–2004")
img3 <- process_image("data/output/ddc_decade_networks/ddc_network_2005-2014.png", "2005–2014")
img4 <- process_image("data/output/ddc_decade_networks/ddc_network_2015-2024.png", "2015–2024")

# Step 2: Append into 2 rows
row1 <- image_append(c(img1, img2))
row2 <- image_append(c(img3, img4))

# Step 3: Stack vertically for 2 × 2 grid
final_image <- image_append(c(row1, row2), stack = TRUE)

# Step 4: Create a title banner image
title_banner <- image_blank(
  width = image_info(final_image)$width,
  height = 100,
  color = "white"
) %>%
  image_annotate(
    "DDC Open Division Player Networks by Decade (3+ Pairings)",
    size = 50,
    gravity = "center",
    color = "black",
    font = "Arial",
    weight = 700
  )

# Step 5: Stack title + grid
full_layout <- image_append(c(title_banner, final_image), stack = TRUE)

# Step 6: Save final image
image_write(full_layout, path = "data/output/ddc_decade_networks/ddc_networks_2x2_full_title.png", format = "png")
```


Ego-centric network
```{r}
library(dplyr)
library(igraph)
library(ggraph)
library(tidygraph)

# Set your target player
target_player <- "John Elsner"

# Filter all pairings involving the player
ego_pairs <- all_teams %>%
  filter(Division == "Open") %>%
  filter(Player1 == target_player | Player2 == target_player) %>%
  mutate(
    teammate = ifelse(Player1 == target_player, Player2, Player1)
  ) %>%
  count(teammate, name = "team_count")

# Build edge list (target → teammate)
edges <- data.frame(
  from = target_player,
  to = ego_pairs$teammate,
  weight = ego_pairs$team_count
)

# Build graph
g <- graph_from_data_frame(edges, directed = FALSE)
tg <- as_tbl_graph(g)

# Plot
ggraph(tg, layout = "star", center = target_player) +
  geom_edge_link(aes(width = weight), color = "gray30", alpha = 0.7) +
  geom_node_point(size = 5, color = ifelse(V(g)$name == target_player, "tomato", "steelblue")) +
  geom_node_text(aes(label = name), repel = TRUE, size = 4) +
  scale_edge_width_continuous(
    breaks = c(1, 3, 5, 10),
    labels = c("1", "3", "5", "10"),
    name = "Teamings"
  ) +
  labs(title = paste("Ego Network for", target_player)) +
  theme_void()
```

Two-layer ego-centric network
```{r}
library(dplyr)
library(igraph)
library(ggraph)
library(tidygraph)

# Ego
target_player <- "Diana Elsner"

# Base pairings (deduplicated)
ddc_pairs <- all_teams %>%
#  filter(Division == "Open") %>%
  mutate(
    player_a = pmin(Player1, Player2),
    player_b = pmax(Player1, Player2)
  ) %>%
  distinct(player_a, player_b)

# Step 1: Ego's direct teammates
ego_edges <- ddc_pairs %>%
  filter(player_a == target_player | player_b == target_player) %>%
  mutate(teammate = ifelse(player_a == target_player, player_b, player_a))

teammates <- unique(ego_edges$teammate)

# Step 2: All pairings where BOTH players are in teammates list (i.e. teammate-to-teammate or teammate-to-others)
second_level_edges <- ddc_pairs %>%
  filter(player_a %in% teammates | player_b %in% teammates) %>%
  filter(player_a != target_player & player_b != target_player)

# Step 3: Combine ego links and second-degree links
all_edges <- bind_rows(
  ego_edges %>% select(from = player_a, to = player_b),
  second_level_edges %>% rename(from = player_a, to = player_b)
) %>%
  filter(!is.na(from), !is.na(to))

# Step 4: Build and plot
g <- graph_from_data_frame(all_edges, directed = FALSE)
tg <- as_tbl_graph(g)

ggraph(tg, layout = "fr") +
  geom_edge_link(color = "gray70", alpha = 0.6) +
  geom_node_point(aes(color = name == target_player), size = 5) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_color_manual(values = c("steelblue", "tomato"), guide = "none") +
  labs(title = paste("2-Level Ego Network for", target_player)) +
  theme_void()

```

Partnership “reach” hierarchy for each player
```{r}
library(dplyr)
library(igraph)
library(tidygraph)

# STEP 1: Build full player network
pair_df <- all_teams %>%
  filter(Division == "Open") %>%
  mutate(
    player_a = pmin(Player1, Player2),
    player_b = pmax(Player1, Player2)
  ) %>%
  distinct(player_a, player_b)

# Create undirected graph
g <- graph_from_data_frame(pair_df, directed = FALSE)
tg <- as_tbl_graph(g)

# STEP 2: For each player, compute 1st, 2nd, 3rd-level neighbors
all_players <- sort(unique(c(pair_df$player_a, pair_df$player_b)))

# Function to compute levels
get_partner_levels <- function(player_name) {
  ego_id <- which(V(g)$name == player_name)
  if (length(ego_id) == 0) return(NULL)

  # Level 1 (direct teammates)
  n1 <- ego(g, order = 1, nodes = ego_id, mode = "all")[[1]] %>% as_ids()
  
  # Level 2 (partners of direct teammates)
  n2 <- ego(g, order = 2, nodes = ego_id, mode = "all")[[1]] %>% as_ids()
  n2 <- setdiff(n2, c(player_name, n1))

  # Level 3 (partners of 2nd-level players)
  n3 <- ego(g, order = 3, nodes = ego_id, mode = "all")[[1]] %>% as_ids()
  n3 <- setdiff(n3, c(player_name, n1, n2))

  data.frame(
    Player = player_name,
    FirstLevel = length(n1),
    SecondLevel = length(n2),
    ThirdLevel = length(n3)
  )
}

# STEP 3: Apply to all players
partner_levels_df <- do.call(rbind, lapply(all_players, get_partner_levels))

# STEP 4: Sort by number of first-level partners
partner_levels_df <- partner_levels_df %>%
  arrange(desc(FirstLevel))

# View the result
head(partner_levels_df)
```

Make the ego graph a Shiny app
```{r}
library(shiny)
library(igraph)
library(ggraph)
library(tidygraph)
library(dplyr)

# Load DDC data
ddc_data <- all_teams

# UI
ui <- fluidPage(
  titlePanel("Double Disc Court Network (ego) Viewer"),

  sidebarLayout(
    sidebarPanel(
      selectInput("division", "Division:", choices = c("Any", "Open", "Women"), selected = "Open"),
      sliderInput("year_range", "Year range:",
                  min = min(ddc_data$Year, na.rm = TRUE),
                  max = max(ddc_data$Year, na.rm = TRUE),
                  value = c(2010, 2024), sep = "", step = 1),
      selectInput("player", "Select a player:", choices = NULL),
      numericInput("level", "Network depth (1 = teammates only, 2 = teammates-of-teammates):",
                   value = 2, min = 1, max = 3)
    ),

    mainPanel(
      plotOutput("egoPlot", height = "700px")
    )
  )
)

# SERVER
server <- function(input, output, session) {

  # Reactive: Filter data by division and year
  filtered_data <- reactive({
    df <- ddc_data %>%
      filter(Year >= input$year_range[1], Year <= input$year_range[2])

    if (input$division != "Any") {
      df <- df %>% filter(Division == input$division)
    }

    df
  })

  # Reactive: update player list based on filtered data
  observe({
  players <- sort(unique(c(filtered_data()$Player1, filtered_data()$Player2)))
  current <- input$player
  if (!is.null(current) && current %in% players) {
    selected <- current
  } else {
    selected <- players[1]
  }
  updateSelectInput(session, "player", choices = players, selected = selected)
})

  # Reactive: Build graph
  graph_data <- reactive({
    df <- filtered_data()

    df %>%
      mutate(
        player_a = pmin(Player1, Player2),
        player_b = pmax(Player1, Player2)
      ) %>%
      distinct(player_a, player_b)
  })

  g <- reactive({
    graph_from_data_frame(graph_data(), directed = FALSE)
  })

  output$egoPlot <- renderPlot({
    req(input$player)
    net <- g()
    ego_id <- which(V(net)$name == input$player)

    if (length(ego_id) == 0) return(NULL)

    # Ego network subgraph
    sub_nodes <- ego(net, order = input$level, nodes = ego_id, mode = "all")[[1]]
    subg <- induced_subgraph(net, vids = sub_nodes)
    tg <- as_tbl_graph(subg)

    # Plot
    ggraph(tg, layout = "fr") +
      geom_edge_link(color = "gray70", alpha = 0.6) +
      geom_node_point(aes(color = name == input$player), size = 5) +
      geom_node_text(aes(label = name), repel = TRUE, size = 3) +
      scale_color_manual(values = c("steelblue", "tomato"), guide = "none") +
      labs(title = paste("Ego Network for", input$player)) +
      theme_void()
  })
}

# Run the app
shinyApp(ui = ui, server = server)
```

For deployment, use the maintained app in `player-team-network/app.R`. Copy an updated `data/output/ddc_team_data.csv` to `player-team-network/ddc_team_data.csv` before deploying.

